---
title: accessibility 1.2.0
author: Daniel Herszenhut
date: '2023-09-21'
slug: accessibility-1-2-0
categories:
  - accessibility
  - package release
tags:
  - accessibility
  - package release
description: "`{accessibility}` provides several functions to calculate accessibility and inequality measures in R. This post overviews the recently released `{accessibility}` v1.2.0."
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`{accessibility}` v1.2.0 has been released. This version includes two new
functions to calculate inequality measures, `concentration_index()` and
`theil_t()`. This post presents an overview of both functions. But first, let's
create an accessibility dataset to use in this demonstration.

```{r}
library(accessibility)

data_dir <- system.file("extdata", package = "accessibility")
travel_matrix <- readRDS(file.path(data_dir, "travel_matrix.rds"))
land_use_data <- readRDS(file.path(data_dir, "land_use_data.rds"))

access <- cumulative_cutoff(
  travel_matrix,
  land_use_data,
  opportunity = "jobs",
  travel_cost = "travel_time",
  cutoff = 30
)
head(access)
```

## Concentration Index

`concentration_index()` calculates the Concentration Index (CI) of a given
accessibility distribution. This measures estimates the extent to which
accessibility inequalities are systematically associated with individuals'
socioeconomic levels. CI values can theoretically vary between -1 and 1 (when
all accessibility is concentrated in the most or in the least disadvantaged
cell, respectively). Negative values indicate that inequalities favor the poor,
while positive values indicate a pro-rich bias.

This function includes an `income` parameter to indicate which variable from the
sociodemographic dataset should be used to rank the population from the least to
the most privileged groups. Any variable that can be used to describe one's
socioeconomic status, such as education level, for example, can be passed to
this argument, as long as it can be numerically ordered (in which higher values
denote higher socioeconomic status).

`concentration_index()` also includes a `type` parameter, used to indicate which
Concentration Index to calculate. This parameter currently supports two values,
`"standard"` and `"corrected"`, which respectively identify the standard
relative CI and the corrected CI, proposed by @erreygers2009correcting.

```{r}
ci <- concentration_index(
  access,
  sociodemographic_data = land_use_data,
  opportunity = "jobs",
  population = "population",
  income = "income_per_capita",
  type = "corrected"
)
ci
```

## Theil T

`theil_t()` calculates the Theil T Index of a given accessibility distribution.
Values range from 0 (when all individuals have exactly the same accessibility
levels) to the natural log of *n*, in which *n* is the number of individuals in
the accessibility dataset.

If the individuals can be classified into mutually exclusive and completely
exhaustive groups (i.e. into groups that do not overlap and cover the entire
population), the index can be decomposed into a between- and a within-groups
inequality component. The function includes a `socioeconomic_groups` parameter
to indicate which variable from the sociodemographic dataset should be used
identify the socioeconomic groups used to calculate these components.

Please note that the output of `theil_t()` varies based on the value of
`socioeconomic_groups`. If `NULL` (the default), the between- and within-groups
components are not calculated, and the function returns a data frame containing
only the total aggregate inequality for the returned area. If
`socioeconomic_groups` is not `NULL`, however, the function returns a list
containing three dataframes: one summarizing the total inequality and the
between- and within-groups components, one listing the contribution of each
group to the between-groups component and another listing the contribution of
each group to the within-groups component. Both behaviors are shown below.

```{r}
theil_without_groups <- theil_t(
  access,
  sociodemographic_data = land_use_data,
  opportunity = "jobs",
  population = "population"
)
theil_without_groups

# some cells are classified as in the decile NA because their income per capita
# is NaN, as they don't have any population. we filter these cells from our
# accessibility data, otherwise the output would include NA values.
# ps: note that subsetting the data like this doesn't affect the assumption that
# groups are completely exhaustive, because cells with NA income decile don't
# have any population.

na_decile_ids <- land_use_data[is.na(land_use_data$income_decile), ]$id
no_na_access <- access[! access$id %in% na_decile_ids, ]
sociodem_data <- land_use_data[! land_use_data$id %in% na_decile_ids, ]

theil_with_groups <- theil_t(
  no_na_access,
  sociodemographic_data = sociodem_data,
  opportunity = "jobs",
  population = "population",
  socioeconomic_groups = "income_decile"
)
theil_with_groups
```

## References
